---
title: "ISA 444: Business Forecasting"
subtitle: "02 - Introduction to Time Series Analysis"
author: Fadel M. Megahed
date: 'Fall 2020'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, xtable)
```

# Preface

### Announcement: Second Speaker Confirmed for Sept. 14, 2020

\begin{figure}
  \centering
  \href{https://www.linkedin.com/in/kyle-cheney-3879a9104/}{\includegraphics[width=0.95\textwidth, height = 0.8\textheight, keepaspectratio, frame]{cheneyLinkedIn}}
\end{figure}


### Quick Refresher based on Last Class

\begin{block}{\textbf{Main Learning Outcomes}}
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Describe course objectives \& structure.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Describe what do we mean by \textcolor{miamired}{forecasting} and explain the \textcolor{miamired}{PIVASE} framework.} \\
			$\quad$ \textcolor{miamired}{\large $\boxtimes$} \textbf{Explain the differences between cross sectional, time series, and panel datasets.} \\
			$\quad$ \textcolor{miamired}{\large $\boxtimes$} \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
\end{block}


### Self-Assessed Skills from Last Class's Mentimeter Poll

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, height=0.77\textheight, keepaspectratio, frame]{Figures/mentimeterClass01a.jpg}
  \caption{Results from Last Class's Mentimeter Poll for Question A}
\end{figure}


### Self-Described Goals for Taking the Class

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth, height=0.77\textheight, keepaspectratio, trim = {0in 19in 0 2in}, clip, frame]{Figures/mentimeterClass01b.jpg}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth, height=0.77\textheight, keepaspectratio, trim = {0in 22in 0 2in}, clip, frame]{Figures/mentimeterClass01c.jpg}
\end{figure}


### Learning Objectives for Today's Class

\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
			\item \textbf{Explain the differences between cross sectional, time series, and panel datasets.} \\
			\item \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
	\end{itemize}
\end{block}




# Types of Data Over Time

### Cross Sectional Data [1]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 1: H1B 2020 Data for Senior Data Scientists at Netflix}}\footnotemark

```{r netflixSalaries, results='asis', echo=FALSE}
if(require(pacman)==FALSE) install.packages(pacman)
pacman::p_load(lubridate, rvest, xtable)
read_html("https://h1bdata.info/index.php?em=NETFLIX+INC&job=SENIOR+DATA+SCIENTIST&city=LOS+GATOS&year=2020") %>%
  html_node("#myTable") %>% html_table() -> netflixSalaries
netflixSalaries$`START DATE` = mdy(netflixSalaries$`START DATE`)
netflixSalaries2020 = netflixSalaries %>%  filter(`START DATE` >= '2020-01-01') %>% 
  select(EMPLOYER, `JOB TITLE`, `BASE SALARY`, LOCATION)
print(xtable(netflixSalaries2020, align = c(rep('c', 5)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data scraped from \url{https://h1bdata.info/index.php?em=NETFLIX+INC&job=SENIOR+DATA+SCIENTIST&city=LOS+GATOS&year=2020} on August 18, 2020 using the \href{https://cran.r-project.org/web/packages/rvest/rvest.pdf}{rvest} package in R. The printing was limited to those individuals who started on/after January 01, 2020, with the filters specified in the URL.}


### Cross Sectional Data [2]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 2: Heart Disease Dataset}}\footnotemark

```{r heart, results='asis', echo=FALSE}
set.seed(444)
heart = read.delim("https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/heart/heart.dat",
                  header = FALSE, sep = " ") %>% select(c(1,2,4,8,14)) %>% sample_n(., 13) 
colnames(heart) = c('age', 'sex', 'restingBP', 'maxHR', 'label')
print(xtable(heart, align = c(rep('c', 6)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data sampled from \href{https://archive.ics.uci.edu/ml/datasets/statlog+(heart)}{this UCI Machine Learning Repository}.}


### Cross Sectional Data [3]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 3: NBA 2019-2020 Leaders - Top 12 in PTS/Game}}\footnotemark

```{r nba, echo = FALSE, results='asis'}
nbaPerGame = read_html("https://www.basketball-reference.com/leagues/NBA_2020_per_game.html") %>% 
  html_node("#per_game_stats") %>% html_table() %>% filter(!PTS %in% "PTS")
nbaPerGame$PTS = as.numeric(nbaPerGame$PTS)
nbaPerGame %>% arrange(desc(PTS)) %>% select(Player, Pos, Age, Tm, G, FG, `FG%`, `eFG%`, PTS) %>% 
  head(n = 12) -> nbaPerGame
print(xtable(nbaPerGame, align = c('l', 'l', rep('c', 8)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data scraped from \href{https://www.basketball-reference.com/leagues/NBA_2020_per_game.html}{Basketball-Reference} on August 18, 2020 using the \href{https://cran.r-project.org/web/packages/rvest/rvest.pdf}{rvest} package in R. The printing with limited to the top 12 players and the selected variables.}



### Time Series Data [1]

**Time Series Data:** Comparable measurements recorded on a single (or a few) variables over time (usually a long period of time).

\textbf{\textcolor{miamired}{Example 1: Net Power Generation in the U.S.}}

\begin{figure}
  \centering
  \href{https://www.eia.gov/electricity/data/browser/}{\includegraphics[width=\textwidth, height=0.5\textheight, keepaspectratio, frame]{eiaChart.png}}
  \caption{Net monthly electricity generation in the U.S. from the \href{https://www.eia.gov/electricity/data/browser/}{U.S. Energy Information Adminstration}.}
\end{figure}


### Time Series Data [2]

\textbf{\textcolor{miamired}{Example 2: Stock prices of U.S. Airlines}}

\vspace{-0.5\baselineskip} 

```{r timeseries2, echo=FALSE, fig.fullwidth = FALSE}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyquant, tidyverse)
airlines = tq_get(c('DAL', 'UAL', 'AAL', 'LUV'),
              from = "2020-01-01")
airlines %>% ggplot(aes(x = date, y = adjusted, group = symbol)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020") +
  facet_wrap(~ symbol, ncol = 2, scale = "free_y") +
  theme_bw()
```

\vspace{-1.25\baselineskip} 

\centering{\textcolor{darkgreen}{As of August 18, 2020: COVID-19 had a substantial impact on these airline stocks.}}



### Panel Data


**Panel Data:** Cross sectional measurements (usually many variable) repeated over time (usually over a few time periods).

\textbf{\textcolor{miamired}{Example: World Bank's Data}}\footnotemark

```{r wb, results='asis', echo=FALSE}
pacman::p_load(wbstats)
macroData = wb_data(country = c('CN','EG','US'),
               indicator = c('NY.GDP.MKTP.KD.ZG', 'SP.DYN.LE00.IN', 'SH.DYN.NMRT', 'SH.HIV.INCD.ZS',
                             'SH.MED.BEDS.ZS', 'SH.MED.PHYS.ZS', 'SH.PRV.SMOK', 'SH.STA.TRAF.P5'),
               mrv = 3, return_wide = TRUE, gapfill = TRUE) %>% 
  select(-c(iso2c, country))
print(xtable(macroData, align = c(rep('c', 11)) ), comment = FALSE, size = '\\tiny', 
      include.rownames=FALSE)
```

\footnotetext{Data queried from the \href{https://datacatalog.worldbank.org/}{World Bank Data Catalog} using the \href{https://cran.r-project.org/web/packages/wbstats/wbstats.pdf}{wbstats} package in R. The printed results show a snapshot of 7 variables (out of a much larger panel dataset). You can think of panel data as a cross-sectional dataset with a longitudinal/time component.}




# Components of a Time Series

### Trend [1]

A **trend** is an increasing or decreasing pattern over time.

```{r trendUP, echo=FALSE}
pacman::p_load(tidyquant, tidyverse)
amazon = tq_get(c('AMZN'), from = "2020-04-01", to = "2020-08-18")
amazon %>% ggplot(aes(x = date, y = adjusted)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020", 
       title = "Increasing Trend", 
       subtitle = "The meteoric rise of $AMZN from 2020-04-01 to 2020-08-18") +
  theme_bw()
```


### Trend [2]

A **trend** is an increasing or decreasing pattern over time.

```{r trendDown, echo=FALSE}
pacman::p_load(tidyquant, tidyverse)
delta = tq_get(c('DAL'), from = "2020-02-03", to = "2020-04-30")
delta %>% ggplot(aes(x = date, y = adjusted)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020", 
       title = "Decreasing Trend", 
       subtitle = "The decline in $DAL from 2020-02-03 to 2020-04-30") +
  theme_bw()
```



### Seasonality [1]

**Seasonality** refers to the property of a time series that displays REGULAR patterns that repeat at a constant frequency (*m*).

\begin{figure}
  \centering 
  \includegraphics[width=\textwidth, height = 0.55\textheight, keepaspectratio]{Figures/seasonal.jpg}
  \caption{A time series with a monthly seasonal pattern.\footnotemark}
\end{figure}

\footnotetext{Figure is from \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}


### Seasonality [2]

**Seasonality** refers to the property of a time series that displays REGULAR patterns that repeat at a constant frequency (*m*).

\vspace{-0.5\baselineskip}

```{r retailSales, echo=FALSE}
pacman::p_load(lubridate)
retail = read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1168&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=RSXFSN&scale=left&cosd=1992-01-01&coed=2020-07-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Monthly&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2020-08-15&revision_date=2020-02-01&nd=2010-01-01")
retail %>% ggplot(aes(x = ymd(DATE), y = RSXFSN)) +
  geom_line() + 
  labs(x=NULL,
       title = "Seasonality with an Additive Trend", 
       subtitle = "Retail (- Food Services) from 2010-01-01 to 2020-02-01",
       caption = 'Data from FRED') +
  theme_bw()
```

### Seasonality [3]

```{r airpassengers, echo=FALSE}
pacman::p_load(ggfortify)
data("AirPassengers")
autoplot(AirPassengers) + 
  labs(x=NULL, title = "Seasonality with a Multiplicative Trend", 
       subtitle = "Non-linear trend & seasonal component grows over time",
       caption = 'AirPassengers R Dataset -- Source: Box, G. E. P., Jenkins, G. M. and \n Reinsel, G. C. (1976) Time Series Analysis, Forecasting and Control.') +
  theme_bw()
```


### Cycle [1]
**Cyclical fluctuations** are somewhat irregular (unknown duration).

\vspace{-0.5\baselineskip}

```{r autoSales, echo=FALSE}
pacman::p_load(lubridate)
autoSales = read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1168&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=TOTALSA&scale=left&cosd=1976-01-01&coed=2020-07-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Monthly&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2020-08-15&revision_date=2020-08-15&nd=1976-01-01")
autoSales %>% ggplot(aes(x = ymd(DATE), y = TOTALSA)) +
  geom_line() + 
  labs(x=NULL,
       title = "The cyclical nature of auto sales",
       caption = 'U.S. Bureau of Economic Analysis, Total Vehicle Sales [TOTALSA], retrieved from FRED, Federal Reserve Bank of St. Louis \n https://fred.stlouisfed.org/series/TOTALSA, August 18, 2020.') +
  theme_bw()
```


### Cycle [2]
**Cyclical fluctuations** are somewhat irregular (unknown duration).

\vspace{-0.5\baselineskip}

\begin{figure}
  \includegraphics[width=0.7\textwidth, height = 0.7\textheight, keepaspectratio]{Figures/Step1.jpg}
  \caption{A gait "cycle" from an IMU attached to the ankle. Joint work with the University at Buffalo.}
\end{figure}



# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
	\begin{itemize}
			\item \textbf{Explain the differences between cross sectional, time series, and panel datasets.}
			\item \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
	\end{itemize}
\end{block}


### Things to Do

 - Thoroughly read Chapter 1 of our book, which can be downloaded from the \href{https://wessexlearning.com/collections/business-analytics/products/principles-of-business-forecasting-2nd-ed}{Publisher} (if you have not gotten your book yet).  
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered. 
 
 - Complete the two graded assignment (refer to next slides or Canvas) for more details.
 

### Graded Assignment 01: Evaluating your Retention/Focus

Please go to \href{https://miamioh.instructure.com/courses/123532/quizzes/317526}{Canvas (click here)} and answer the four questions. **Due August 24, 2020 [2:51 PM,
Ohio local time].** 

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of the material covered in Class 01. In order to prepare for this, you should have either actively attended class and/or watched the recording from WebEx.

**General Guidelines:**  

\vspace{-0.5\baselineskip}

  - Individual assignment.  
  - This is a timed assignment (i.e. once you start the assignment you will have 25 minutes to complete 4 questions). The purpose of the time limit is to help you evaluate how much you have retained/understood from class. If the concepts we covered today are well-understood, this should take 10-15 minutes.  
  - Proctorio is NOT required for this assignment.  
  - You will need to have R installed (or accessible through the \href{https://virtualpc.fsb.miamioh.edu/RDWeb/Pages/en-US/default.aspx}{Remote Desktop})


### Graded Assignment 02: Evaluating your Understanding

Please go to \href{https://miamioh.instructure.com/courses/123532/quizzes/320286}{Canvas (click here)} and answer the four questions. **Due August 24, 2020 [2:51 PM,
Ohio local time].** 

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of the material covered up to the End of Class 02. In order to prepare for this, you should have either actively attended class and/or watched the recording from WebEx. Furthermore, you should have thoroughly read [Chapter 01 of the book](https://wessexlearning.com/collections/business-analytics/products/principles-of-business-forecasting-2nd-ed).

**General Guidelines:**  

\vspace{-0.5\baselineskip}

  - Individual assignment.  
  - This is **NOT** a timed assignment. If the concepts we covered are well-understood, this should take 20-35 minutes.  
  - Proctorio is NOT required for this assignment.  
  - You will need to have R installed (or accessible through the \href{https://virtualpc.fsb.miamioh.edu/RDWeb/Pages/en-US/default.aspx}{Remote Desktop})


---

\maketitle