---
title: "ISA 444: Business Forecasting"
subtitle: "02 - Introduction to Time Series Analysis"
author: Fadel M. Megahed
date: 'Spring 2021'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, xtable)
url_seasonal = 'https://cdnx.numxl.com/storage/originals/f0/50/7593920758_1a84d6f191_z.jpg'
```

# Preface

### Quick Refresher based on Last Class

\begin{block}{\textbf{Main Learning Outcomes}}
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Describe course objectives \& structure.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Describe what do we mean by \textcolor{miamired}{forecasting}} \\
			$\quad$ \textcolor{miamired}{\large $\boxtimes$} \textbf{Explain the \textcolor{miamired}{PIVASE} framework} \\
\end{block}


### Self-Assessed Skills from Last Class's Mentimeter Poll

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, height=0.77\textheight, keepaspectratio, frame]{Figures/mentimeterClass01a.jpg}
  \caption{Results from Last Class's Mentimeter Poll for Question A}
\end{figure}


### Self-Described Goals for Taking the Class

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth, height=0.77\textheight, keepaspectratio, trim = {0in 19in 0 2in}, clip, frame]{Figures/mentimeterClass01b.jpg}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth, height=0.77\textheight, keepaspectratio, trim = {0in 22in 0 2in}, clip, frame]{Figures/mentimeterClass01c.jpg}
\end{figure}


### Learning Objectives for Today's Class

\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
    \item \textbf{Explain the \textcolor{miamired}{PIVASE} framework.} \\
			\item \textbf{Explain the differences between cross sectional, time series, and panel datasets.} \\
			\item \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
	\end{itemize}
\end{block}


# Why do we Forecast?

### The COVID-19 Activity

Based on our discussion last class, I have asked to use the [COVID-19 package](https://covid19datahub.io/articles/api/r.html) to plot the confirmed COVID-19 cases for the states of Ohio and Kentucky. You should have created two plots: (a) plot containing both of the states as two seperate lines in a line plot, and (b) a paneled plot similar to the output depicted in Slide 21 in [Slide Deck 01](https://miamioh.instructure.com/courses/142177/files/18795137?module_item_id=2857229). **I will quickly walk you through the solution to help ensure that the concepts discussed in last class are well understood.**

**Hints:**  

  (1) Once you have installed the package inspect the `covid19` function using the command `?covid19()`.  
  (2) Set the country argument to the US.   
  (3) Use an apporpriate value for the level argument based on `?covid19()`.  
  (4) Set the starting date to '2020-03-01'
  (5) Filter the obtained data to only include data for the two states.


### Why do we Forecast? -- A Conceptual Framework (PIVASE)\footnotemark

- **Purpose:** What do we hope to achieve by generating the forecast? That is, what plans are dependent upon the results of the forecasting exercise? How far ahead do we wish to forecast? We refer to this period as the forecasting horizon.  

- **Information:** What do we know that may help us in forecasting. And when will we know it? Detailed data is only useful if it is available in timely fashion.   

- **Value:** How valuable is the forecast? What would you pay for perfect knowledge?   

- **Analysis:** From analyzing the data can we develop a model that captures its characteristics? And how does it perform on new (hold-out sample) data?   

- **System:** What models and software are needed to meet the needs of the organization?  

- **Evaluation:** How do we know whether a particular forecasting exercise was effective and what the potential is for improvement?

\footnotetext{From: Ord, K., Fildes, R., \& Kourentzes, N. (2017). Principles of Business Forecasting (2nd ed., p. 3-6).}


### Why do Businesses Forecast?

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{Figures/businessForecasts.png}
  \caption{Some of the typical forecasting needs of many organizations.\footnotemark}
\end{figure}

\footnotetext{From: Ord, K., Fildes, R., \& Kourentzes, N. (2017). Principles of Business Forecasting (2nd ed., p. 7).}



# Types of Data Over Time

### Cross Sectional Data [1]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 1: H1B 2020 Data for Senior Data Scientists at Netflix}}\footnotemark

```{r netflixSalaries, results='asis', echo=FALSE}
if(require(pacman)==FALSE) install.packages(pacman)
pacman::p_load(lubridate, rvest, xtable)
read_html("https://h1bdata.info/index.php?em=NETFLIX+INC&job=SENIOR+DATA+SCIENTIST&city=LOS+GATOS&year=2020") %>%
  html_node("#myTable") %>% html_table() -> netflixSalaries
netflixSalaries$`START DATE` = mdy(netflixSalaries$`START DATE`)
netflixSalaries2020 = netflixSalaries %>%  filter(`START DATE` >= '2020-01-01') %>% 
  select(EMPLOYER, `JOB TITLE`, `BASE SALARY`, LOCATION)
print(xtable(netflixSalaries2020, align = c(rep('c', 5)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data scraped from \url{https://h1bdata.info/index.php?em=NETFLIX+INC&job=SENIOR+DATA+SCIENTIST&city=LOS+GATOS&year=2020} on January 26, 2021 using the \href{https://cran.r-project.org/web/packages/rvest/rvest.pdf}{rvest} package in R. The printing was limited to those individuals who started on/after January 01, 2020, with the filters specified in the URL.}


### Cross Sectional Data [2]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 2: Heart Disease Dataset}}\footnotemark

```{r heart, results='asis', echo=FALSE}
set.seed(444)
heart = read.delim("https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/heart/heart.dat",
                  header = FALSE, sep = " ") %>% select(c(1,2,4,8,14)) %>% sample_n(., 13) 
colnames(heart) = c('age', 'sex', 'restingBP', 'maxHR', 'label')
print(xtable(heart, align = c(rep('c', 6)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data sampled from \href{https://archive.ics.uci.edu/ml/datasets/statlog+(heart)}{this UCI Machine Learning Repository}.}


### Cross Sectional Data [3]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 3: NBA 2020-2021 Leaders - Top 12 in PTS/Game}}\footnotemark

```{r nba, echo = FALSE, results='asis'}
nbaPerGame = read_html("https://www.basketball-reference.com/leagues/NBA_2021_per_game.html") %>% 
  html_node("#per_game_stats") %>% html_table() %>% filter(!PTS %in% "PTS")
nbaPerGame$PTS = as.numeric(nbaPerGame$PTS)
nbaPerGame %>% arrange(desc(PTS)) %>% select(Player, Pos, Age, Tm, G, FG, `FG%`, `eFG%`, PTS) %>% 
  head(n = 12) -> nbaPerGame
print(xtable(nbaPerGame, align = c('l', 'l', rep('c', 8)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data scraped from \href{https://www.basketball-reference.com/leagues/NBA_2020_per_game.html}{Basketball-Reference} on January 26, 2021 using the \href{https://cran.r-project.org/web/packages/rvest/rvest.pdf}{rvest} package in R. The printing with limited to the top 12 players and the selected variables.}



### Time Series Data [1]

**Time Series Data:** Comparable measurements recorded on a single (or a few) variables over time (usually a long period of time).

\textbf{\textcolor{miamired}{Example 1: Net Power Generation in the U.S.}}

\begin{figure}
  \centering
  \href{https://www.eia.gov/electricity/data/browser/}{\includegraphics[width=\textwidth, height=0.5\textheight, keepaspectratio, frame]{eiaChart.png}}
  \caption{Net monthly electricity generation in the U.S. from the \href{https://www.eia.gov/electricity/data/browser/}{U.S. Energy Information Adminstration}.}
\end{figure}


### Time Series Data [2]

\textbf{\textcolor{miamired}{Example 2: Stock prices of U.S. Airlines}}

\vspace{-0.5\baselineskip} 

```{r timeseries2, echo=FALSE, fig.fullwidth = FALSE}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyquant, tidyverse, scales)
airlines = tq_get(c('DAL', 'UAL', 'AAL', 'LUV'),
              from = "2020-01-01")
airlines %>% ggplot(aes(x = date, y = adjusted, group = symbol)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020") +
  facet_wrap(~ symbol, ncol = 2, scale = "free_y") +
  scale_x_date(breaks = pretty_breaks(n=4)) +
  theme_bw()
```

\vspace{-1.25\baselineskip} 

\centering{\textcolor{miamired}{As of January, 26 2021: COVID-19 had a substantial impact on these airline stocks.}}



### Panel Data


**Panel Data:** Cross sectional measurements (usually many variables) repeated over time (usually over a few time periods).

\textbf{\textcolor{miamired}{Example: World Bank's Data}}\footnotemark

```{r wb, results='asis', echo=FALSE}
pacman::p_load(wbstats)
macroData = wb_data(country = c('CN','EG','US'),
               indicator = c('NY.GDP.MKTP.KD.ZG', 'SP.DYN.LE00.IN', 'SH.DYN.NMRT', 'SH.HIV.INCD.ZS',
                             'SH.MED.BEDS.ZS', 'SH.MED.PHYS.ZS', 'SH.PRV.SMOK', 'SH.STA.TRAF.P5'),
               mrv = 3, return_wide = TRUE, gapfill = TRUE) %>% 
  select(-c(iso2c, country))
print(xtable(macroData, align = c(rep('c', 11)) ), comment = FALSE, size = '\\tiny', 
      include.rownames=FALSE)
```

\footnotetext{Data queried from the \href{https://datacatalog.worldbank.org/}{World Bank Data Catalog} using the \href{https://cran.r-project.org/web/packages/wbstats/wbstats.pdf}{wbstats} package in R. The printed results show a snapshot of 7 variables (out of a much larger panel dataset). You can think of panel data as a cross-sectional dataset with a longitudinal/time component.}




# Components of a Time Series

### Trend [1]

A **trend** is an increasing or decreasing pattern over time.

```{r trendUP, echo=FALSE}
pacman::p_load(tidyquant, tidyverse)
amazon = tq_get(c('AMZN'), from = "2020-04-01", to = "2020-08-01")
amazon %>% ggplot(aes(x = date, y = adjusted)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020", 
       title = "Increasing Trend", 
       subtitle = "The meteoric rise of $AMZN from 2020-04-01 to 2020-08-01") +
  theme_bw()
```


### Trend [2]

A **trend** is an increasing or decreasing pattern over time.

```{r trendDown, echo=FALSE}
pacman::p_load(tidyquant, tidyverse)
delta = tq_get(c('DAL'), from = "2020-02-03", to = "2020-04-30")
delta %>% ggplot(aes(x = date, y = adjusted)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020", 
       title = "Decreasing Trend", 
       subtitle = "The decline in $DAL from 2020-02-03 to 2020-04-30") +
  theme_bw()
```



### Seasonality [1]

**Seasonality** refers to the property of a time series that displays REGULAR patterns that repeat at a constant frequency (*m*).

\begin{figure}
  \centering 
  \includegraphics[width=\textwidth, height = 0.55\textheight, keepaspectratio]{Figures/seasonal.jpg}
  \caption{A time series with a monthly seasonal pattern.\footnotemark}
\end{figure}

\footnotetext{Figure is from \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}


### Seasonality [2]

**Seasonality** refers to the property of a time series that displays REGULAR patterns that repeat at a constant frequency (*m*).

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, height=0.55\textheight, keepaspectratio, frame]{Figures/seasonalAdditive.jpg}
  \caption{Seasonality with an Additive Trend\footnotemark.}
\end{figure}

\footnotetext{Figure is from \href{https://support.numxl.com/hc/en-us/articles/115000149703-Patterns-Unplugged}{Patterns Unplugged - Help Center} published online on December 28, 2016.}


### Seasonality [3]

```{r airpassengers, echo=FALSE}
pacman::p_load(ggfortify)
data("AirPassengers")
autoplot(AirPassengers) + 
  labs(x=NULL, title = "Seasonality with a Multiplicative Trend", 
       subtitle = "Non-linear trend & seasonal component grows over time",
       caption = 'AirPassengers R Dataset -- Source: Box, G. E. P., Jenkins, G. M. and \n Reinsel, G. C. (1976) Time Series Analysis, Forecasting and Control.') +
  theme_bw()
```


### Cycle [1]
**Cyclical fluctuations** are somewhat irregular (unknown duration).

\vspace{-0.5\baselineskip}

```{r autoSales, echo=FALSE}
pacman::p_load(lubridate)
autoSales = read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1168&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=TOTALSA&scale=left&cosd=1976-01-01&coed=2020-11-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Monthly&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2021-01-26&revision_date=2021-01-26&nd=1976-01-01")
autoSales %>% ggplot(aes(x = ymd(DATE), y = TOTALSA)) +
  geom_line() + 
  labs(x=NULL,
       title = "The cyclical nature of auto sales",
       caption = 'U.S. Bureau of Economic Analysis, Total Vehicle Sales [TOTALSA], retrieved from FRED, Federal Reserve Bank of St. Louis \n https://fred.stlouisfed.org/series/TOTALSA, January 26, 2021.') +
  theme_bw()
```


### Cycle [2]
**Cyclical fluctuations** are somewhat irregular (unknown duration).

\vspace{-0.5\baselineskip}

\begin{figure}
  \includegraphics[width=0.7\textwidth, height = 0.7\textheight, keepaspectratio]{Figures/Step1.jpg}
  \caption{A gait "cycle" from an IMU attached to the ankle. Joint work with the University at Buffalo.}
\end{figure}



# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
	\begin{itemize}
			\item \textbf{Explain the differences between cross sectional, time series, and panel datasets.}
			\item \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
	\end{itemize}
\end{block}


### Things to Do

 - Thoroughly read Chapter 1 of our book, which can be downloaded from the \href{https://wessexlearning.com/collections/business-analytics/products/principles-of-business-forecasting-2nd-ed}{Publisher} (if you have not gotten your book yet).  
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered.  
 - **Optional, but highly recommended,** go through the [self-paced interactive review for Week 01](http://rstudio.fsb.miamioh.edu:3838/megahefm/isa444/week01/).
 
 - **Required:** Complete the graded assignment.
 

### Self-Paced Interactive Review for Week 01

To help you brush your \texttt{R} skills and assess your understanding of the material covered in Week 01, I have created a self-paced interactive review. The review can be accessed at <http://rstudio.fsb.miamioh.edu:3838/megahefm/isa444/week01/>. 


This review is optional, but if you plan on completing it, you are advised to do the following prior to going through the self-paced review.  

  1. Read Chapter 01 from our textbook.   
  2. Have completed the in-class examples in the lecture notes for Week 01.  
  3. Be able to dedicate about 20-40 minutes to go through this entire tutorial.
  
  
You are encouraged to complete this review prior to our first Week 02 class; there is no submission associated with this review.


### Graded Assignment 02: Evaluating your Understanding

Please go to \href{https://miamioh.instructure.com/courses/142177/quizzes/367195}{Canvas (click here)} and answer the four questions. **Due February 01, 2021 [11:40 AM, Ohio local time].** 

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of the material covered up to the End of Class 02. In order to prepare for this, you should have either actively attended class and/or watched the recording from WebEx. Furthermore, you should have thoroughly read [Chapter 01 of the book](https://wessexlearning.com/collections/business-analytics/products/principles-of-business-forecasting-2nd-ed).

**General Guidelines:**  

\vspace{-0.5\baselineskip}

  - Individual assignment.  
  - This is **NOT** a timed assignment. If the concepts we covered are well-understood, this should take 20-35 minutes.  
  - Proctorio is NOT required for this assignment.  
  - You will need to have R installed (or accessible through the \href{https://virtualpc.fsb.miamioh.edu/RDWeb/Pages/en-US/default.aspx}{Remote Desktop})


---

\maketitle