---
title: "Class 28: Communicating Via Flexdashboards"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
runtime: shiny
---

```{r setup, include = FALSE}
pacman::p_load(tidyverse, tidyquant, magrittr, fpp2, plotly, dygraphs, 
               DT, imputeTS, fontawesome, flexdashboard, shiny)

# Currency formating based on https://stackoverflow.com/a/56939122/10156153
mycurrency <- function(x){
  return(paste("$", formatC(as.numeric(x), format="f", digits=2, big.mark=",")))
}

imageURL = 'https://pbs.twimg.com/profile_images/422580700290613248/HDNDEppR_400x400.png'
```
  
```{r global, include=FALSE, cache=TRUE, cache.extra = Sys.Date()}
tickers = c('AAPL', 'CSCO', 'DAL', 'GOLD', 'JNJ', 'IQV', 'MCD', 'PYPL', 'TMUS', 'DHR', 'MDT',
           'UNH', 'TSLA', 'F', 'AMZN', 'WMT', 'NFLX', 'GOOG', 'TWTR', 'SBUX', 'HD', 'PG',
           'DS', 'CI', 'MDLZ')
```

# Column {.sidebar}

**Global Choices:**

```{r globalChoices}
selectInput("periodicity", label = "Unit of Analysis:",
            choices = c("daily", "weekly", "monthly"), selected = "monthly")

selectInput(inputId = "tickers", label = "Stock Tickers: (Multiple)", 
               choices = c('AAPL', 'CSCO', 'DAL', 'GOLD', 'JNJ', 'IQV', 'MCD', 'PYPL', 'TMUS', 'DHR', 'MDT',
           'UNH', 'TSLA', 'FB', 'AMZN', 'WMT', 'NFLX', 'GOOG', 'TWTR', 'SBUX', 'HD', 'PG',
           'DS', 'CI', 'MDLZ'), 
           selected = c("FB", "AAPL", "AMZN", "NFLX"), multiple = TRUE) ## and switch multiple to True

numericInput(inputId = "panels", label = "Select Number of Columns for Plot", 
             min = 1, max = 4, step = 1, value = 2) ## and switch multiple to True
```


---

<br> <br><br><br><br>

**Filters for Candle Chart:**
```{r filterCandleChart}
dateInput("date1", "Starting Date:", value = "2016-01-01")
dateInput("date2", "Ending Date:", value = Sys.Date()-1)
```


---

<br> <br><br><br> <br>


**Created By:** Fadel Megahed

**Version:** 0.1.0

**Last Updated:** `r Sys.Date() %>% format("%b %d, %Y")`

---

**Contact Information:**

`r fa("twitter", fill = "steelblue")` $~$  [\@FadelMegahed](https://twitter.com/FadelMegahed)

`r fa("linkedin", fill = "steelblue")` $~$  [Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4/)

![](`r imageURL`){width=200px}


# Getting the Data

## Column

### Extracting the Data

```{r rawData}
stocks = reactive({
  tq_get(x = input$tickers, from = '2010-01-01', to = Sys.Date() - 1, periodicity = input$periodicity)
})

DT::renderDataTable(
  datatable(
    stocks() %>% select(symbol, date, adjusted, volume), 
    filter = 'top',
    extensions = c('Buttons'), 
    options = list(
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
      pageLength = 25,
      rownames = F,
      digits = c(0, 0, 0, 2, 0)
    ) 
  ) %>% 
    formatCurrency('adjusted') %>% formatCurrency('volume', '', digits = 0)
)
```



# Visualizations

## Row

### Plotting the Adjusted Closing Prices
```{r tsGraph}
renderPlotly({
  p1 = stocks() %>% 
    ggplot(aes(x = date, y = adjusted, group = symbol, color = symbol, 
               text = paste('Symbol:', symbol,
                            '<br>Date:', as.Date(date),
                            '<br>Adjusted:', mycurrency(adjusted) ) 
               )
           ) +
    geom_line(size = 1.5) + theme_bw(base_size = 18) +
    scale_x_date(breaks = scales::pretty_breaks(n= 6)) +
    scale_color_viridis_d() + 
    scale_y_continuous(labels = scales::dollar_format()) +
    facet_wrap(~ symbol, ncol = input$panels, scales = "free_y") +
    labs(x = '') +
    theme(legend.position = "none")
  
  ggplotly(p1, tooltip = c("text")) 
})
```

## Row

### Panel Candlestick Plot (Static)
  
```{r tsCandleStick}
dateInput("date1a", "Starting Date:", value = "2016-01-01")
dateInput("date2a", "Ending Date:", value = Sys.Date()-1)
renderPlot({
  stocks() %>% 
    filter(date >= input$date1 & date <= input$date2 +1) %>% 
    ggplot(aes(x = date, y = close, group = symbol)) +
    geom_candlestick(aes(open = open, high = high, low = low, close = close), size = 3, 
                     fill_up = 'darkgreen', colour_up = 'darkgreen') +
    labs(y = "Closing Price", x = "") +
    theme_bw(base_size = 18) +
    scale_x_date(breaks = scales::pretty_breaks(n= 6)) +
    scale_y_continuous(labels = scales::dollar_format()) +
    facet_wrap(~ symbol, ncol = input$panels, scales = "free_y")
})
```


# Forecasting

For the purpose of class time, we will only deploy a naive forecast. **Obviously, you can try to include some of the more advanced models discussed in class. Perhaps, include them as a choice for the user to pick which model to run.**

## Row

### Naive Predictions

```{r naiveFC}
df =reactive({ 
  temp = stocks()
  temp %<>% group_by(symbol) 
  temp %<>% mutate(naiveFC = lag(adjusted)) 
})

DT::renderDataTable(
  datatable(
    df() %>% select(symbol, date, adjusted, naiveFC), 
    filter = 'top',
    extensions = c('Buttons'), 
    options = list(
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
      pageLength = 10,
      rownames = F,
      digits = c(0, 0, 0, 2, 2)
    ) 
  ) %>% 
    formatCurrency(c('adjusted', 'naiveFC') )
)

```

## Row

### Actual vs Fitted

```{r actualFitted}
renderPlotly({
  p3 = df() %>% 
    ggplot(aes(x = date, group = symbol, color = symbol, 
               text = paste('Symbol:', symbol,
                            '<br>Date:', as.Date(date),
                            '<br>Adjusted:', mycurrency(adjusted),
                            '<br>Fitted:', mycurrency(naiveFC)) 
               )
           ) +
    geom_line(aes(y = adjusted, color = 'adjusted'), size = 1) + 
    geom_line(aes(y = naiveFC, color = 'naiveFC'), size = 1) +
    theme_bw(base_size = 12) +
    scale_x_date(breaks = scales::pretty_breaks(n= 6)) +
    scale_color_manual(values = c('adjusted' = 'black', 'naiveFC' = 'gray')) + 
    scale_y_continuous(labels = scales::dollar_format()) +
    facet_wrap(~ symbol, ncol = input$panels, scales = "free_y") +
    theme(legend.position = "top")
  
  ggplotly(p3, tooltip = c("text")) 
})

```

### Forecast Accuracy Summary

Below, we print the error metrics for the `naive forecast`.

```{r naiveFCAccuracy}
acc = reactive({
  temp = df() %>% group_by(symbol)
  
  temp %<>% mutate(error = adjusted - naiveFC,
                   SE = error^2, 
                   AE = abs(error), 
                   PE = 100*error/adjusted,
                   APE = abs(PE))
  
  accMetrics = temp %>% 
    summarise(ME = mean(error, na.rm = T),
              RMSE = sqrt(mean(SE, na.rm = T)),
              MAE = mean(AE, na.rm = T),
              MPE = mean(PE, na.rm = T),
              MAPE = mean(APE, na.rm = T)
              )
  
  accMetrics %>%  mutate_if(is.numeric, round, digits = 2) 
})

renderDataTable(
  datatable(
    acc(),
    options = list(rownames = F)
    )
)

```