---
title: "ISA 444: Business Forecasting"
subtitle: "21 - ARIMA Models (Cont.)"
author: Fadel M. Megahed
date: 'Fall 2020'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.8,
                      fig.height= 2.9,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, tidyquant, xtable, magrittr, fpp2, ggpubr)
```

# Preface

### Recap of What we Have Covered Last Two Weeks

**ARIMA Models:** Models we considered may have three components, an autoregressive component (AR), integrated (I for differencing) and a moving average component (MA).

\begin{block}{\textbf{Main Learning Outcomes from Last 3 Classes}}
  \begin{itemize}
			 \item \textbf{Describe the behavior of the ACF and PACF of an ARMA (p,q) process.}
			\item \textbf{Fit an ARMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARIMA model.}
			 \item \textbf{Use nonseasonal differencing to attain stationarity for a time series.}
			\item \textbf{Fit an ARIMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARIMA model.}
			 \item \textbf{Show that you can fit reasonable ARIMA models based on both simulated and actual data cases.}
	\end{itemize}
\end{block}


### Learning Outcomes for Today's Class
\begin{block}{\textbf{Main Learning Outcomes\footnotemark}}
  \begin{itemize}
      \item \textbf{Describe AIC, AICc, and BIC and how they are used to measure model fit.}
      \item \textbf{Describe the algorithm used within the auto.arima() function to fit an ARIMA model.}
      \item \textbf{Describe the results of the auto.arima() function.}
	\end{itemize}
\end{block}

\footnotetext{Today's class slides are based on  \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}

# Some Measures of Goodness of Fit

### Additional Measures of Goodness of Fit {.allowframebreaks}

**Akaike’s Information Criterion:** Akaike suggests measuring the goodness of fit of a model by balancing the error of the fit against the number of parameters in the model.

$AIC = 2k - 2\ln{\hat{(L)}}$, 

where $k$ is the number of parameters in the model and $\hat{L}$ is the sample likelihood function. The value of $k$ that gives a minimum AIC gives the best model. This is simply a penalty imposed on the error variance for the number of
parameters in the model.


**Bias Corrected AIC:** $AICc = AIC + \frac{2k^2 + 2k}{n - k - 1}$.

The AICc is usually preferred over the AIC.


**Bayesian Information Criterion:** $BIC = \ln{(n)}k - 2\ln{\hat{(L)}}$. 

BIC is also known as the Schwarz Information Criterion (SIC). The BIC has
a larger penalty for model size and tends to choose smaller models.


**Studies have shown:**  

  - BIC does well at getting correct model in large samples.  
  - AICc tends to get correct models in smaller samples with a large number of parameters.
  
  
**Why did we discuss these metrics today?**  

  - They were printed with some of the models that we have examined in class.   
  - They are used with the `auto.arima()`, which comes from the [forecast package](https://cran.r-project.org/web/packages/forecast/forecast.pdf) (loaded with `fpp2`).


# The `auto.arima()` Function

### The `auto.arima()` Function {.allowframebreaks}

The auto.arima() function can be used to automatically fit ARIMA models to a time series. It is a useful function, but it should be used with caution.

**The function**  

  - Uses “brute force” to fit many models and then selects the “best” based on a certain model criterion  
  - Works best when the data are stationary, but can be used with nonstationary data  
  - Tends to overfit the data  
  - Should always be used as a starting point for selecting a model and all models derived from the `auto.arima()` function should be properly vetted and evaluated.


The `auto.arima()` function combines  
  - Unit root tests (KPSS by default)  
  - Minimization of AICc
to obtain an $ARIMA(p,d,q)$ model using the following algorithm:


  1. Determine the number of differences, $d$, using a sequence of KPSS tests.  
  
  2. Determine $p$ and $q$ by minimizing AICc after differencing the data d times. Rather than considering all possible $p$ and $q$ combinations, a stepwise approach is taken.  
      - The best initial model with lowest AICc is selected from the following four:   
        - ARIMA (2,$d$,2),  
        - ARIMA (0,$d$,0),  
        - ARIMA (1,$d$,0), and   
        - ARIMA (1,$d$,0).  
        - *If d=0, then a constant, $c$, is included. If $d \ge 1$, then the constant is set to 0. The results of this step is called the current model.*  
      - Variations on the current model are considered by  
        - Vary $p$ and/or $q$ from current model by $\pm 1$  
        - Include/exclude $c$ from current model.  
        - The best model considered so far (either current or one of variations) becomes the *new current model*. 
      - Repeat previous step until no lower AICc can be found.
    

### Live Coding: Example 1

In class, we will use a different snapshot of the GNP data that we have explored in class so far. The purpose of this different snapshot is two-fold:  
  - we are not sure whether the model we fit last class would be appropriate (so this is somewhat of a revision of what to do in order to fit the "best" ARIMA model by hand).      - Walk you through the process of finding the model selected from the `auto.arima()`
  
```{r autoArima}
pacman::p_load(astsa)
gnpData = gnp # will be loaded from the astsa package until 2002

# We will build on this example in class
```


### Live Coding: Example 2
```{r birthData}
birthData = birth # also from the astsa package
```




### Things to Do to Prepare for Next Class

 - Thoroughly read Chapters 6.2-6.8 of our textbook.
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered.  
 
 


---

\maketitle