---
title: "Communicating your Results to Senior Management"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

pacman::p_load(tidyverse, magrittr, tidyquant, # our typical packages
               plotly, # for interactive visuals
               shiny, flexdashboard, DT, # for the shiny/flexdashboard functions
               fontawesome) # for neat looking icons

tickers = c('AAPL', 'AMZN', 'NFLX', 'FB' , 'GOOG', 'TWTR',
            'F', 'TSLA', 'GE', 'C',
            'MCD', 'WMT', 
            'DAL', 'AA')
```


# Sidebar {.sidebar}

**Global Choices:**

```{r globalChoices}
selectInput(inputId = 'periodicity', label = 'Select Unit of Aggregation',
            choices = c('daily', 'weekly', 'monthly'),
            selected = 'monthly')

selectInput(inputId = 'tickers', label = 'Select Multiple Stocks', 
            choices = tickers, 
            selected = c('AAPL', 'AMZN', 'FB', 'NFLX'), 
            multiple = T)


```

<br><br>

---

<br><br>

**CandleStick Filter:**
```{r candlestickFilter}
dateInput(inputId = 'start', label = 'Starting Date', value = Sys.Date() - months(13))

```

<br><br>

---

<br><br>

**Created by:** Fadel Megahed

**Version:** 0.0.1

**Last Updated:** `r format(Sys.Date(), format = "%b %d, %Y")`

<br><br>

---

<br><br>

**Contact Information:**

`r fa("twitter", fill = "steelblue")` $~$ [\@FadelMegahed](https://twitter.com/FadelMegahed)

`r fa("linkedin", fill = "steelblue")` $~$ [Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4/)






# Getting the Data

## Row

### Extracting the Data

```{r rawData}
stocks = reactive({
  tq_get(x = input$tickers, from = '2010-01-01', to = Sys.Date() -1, 
         periodicity = input$periodicity)
})


renderDataTable(
  datatable(
    data = stocks() %>% select(symbol, date, adjusted, volume),
    extensions = 'Buttons', 
    options = list(
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
      pageLength = 20
    ) 
  ) %>% formatCurrency('adjusted') %>% formatCurrency('volume', currency = '', digits = 0)
)

```


# Data Visualization

## Row

### Plotly of the TS

```{r timeSeriesPlot}
renderPlotly({
  p1 = stocks() %>% 
    ggplot(aes(x = date, y = adjusted, group = symbol, color = symbol,
               text = paste('Symbol:', symbol,
                            '<br>Date:', date, 
                            '<br>Adjusted', round(adjusted, digits = 2) )
    ) 
    ) +
    geom_line() +
    scale_x_date(breaks = scales::pretty_breaks(n = 5)) +
    facet_wrap(~symbol, ncol = 2, scales = 'free_y') + theme_bw() +
    theme(legend.position = 'none')

ggplotly(p1, tooltip = "text")
})
```

## Row

### Candlestick Chart
```{r candlestick}
renderPlot({
  stocks() %>% 
    filter(date >= input$start) %>% 
    ggplot(aes(x = date, y = adjusted, group = symbol)) +
    scale_x_date(breaks = scales::pretty_breaks(n= 12)) +
    geom_candlestick(aes(open = open, high = high, low = low, close = close), 
    fill_up = 'darkgreen' , colour_up = 'darkgreen', size = 3) +
    facet_wrap(~symbol, scales = 'free_y') + theme_bw()
})

```


# Time Series Forecasting

**Note:** We will explore the most simple forecast today to make our lives easier. We will only look at the **naive forecast.** Obviously, you may/should explore other approaches.

## Row

### Naive Forecast

```{r naiveF}
df = reactive({
  dfN = stocks() %>% group_by(symbol)
  dfN %<>% mutate(naiveFC = lag(adjusted)) 
})

renderDataTable({
  datatable(
  df() %>% select(symbol, date, adjusted, naiveFC),
    extensions = 'Buttons', 
    options = list(
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
      pageLength = 20
    ) 
  ) %>% formatCurrency('adjusted') %>% formatCurrency('naiveFC')
})
```


## Row

### Actual Vs Fitted

```{r actualFitted}
renderPlotly({
  p2 = df() %>% 
    ggplot(aes(x = date, group = symbol, color = symbol,
               text = paste('Symbol:', symbol,
                            '<br>Date:', date, 
                            '<br>Adjusted', round(adjusted, digits = 2) )
    ) 
    ) +
    geom_line(aes(y = adjusted, color = 'adjusted')) +
    geom_line(aes(y = naiveFC, color = 'nForecast')) +
    scale_x_date(breaks = scales::pretty_breaks(n = 5)) +
    scale_color_manual(values = c('adjusted' = 'black', 'nForecast' = 'gray')) +
    facet_wrap(~symbol, ncol = 2, scales = 'free_y') + theme_bw()

ggplotly(p2, tooltip = "text")
})
```


### Forecast Accuracy (Per Symbol)

```{r forecastAcc}
accuracy = reactive({
  temp = df() %>% 
    group_by(symbol) # we do not really need this since it was grouped before
  
  temp %<>% mutate(error = adjusted - naiveFC,
                   SE = error^2, 
                   AE = abs(error),
                   PE = 100*error/adjusted,
                   APE = abs(PE))
  
  metrics = temp %>% 
    summarise(ME = mean(error, na.rm = T), 
              RMSE = mean(SE, na.rm = T) %>% sqrt(),
              MAE = mean(AE, na.rm = T),
              MPE = mean(PE, na.rm = T),
              MAPE = mean(APE, na.rm = T)) %>% 
    mutate_if(is.numeric, round, digits = 2)
})

renderDataTable({
  datatable(
    accuracy()
    )
})

```