---
title: "ISA 444: Business Forecasting"
subtitle: "20 - Seasonal ARIMA Models"
author: Fadel M. Megahed
date: 'Spring 2021'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.8,
                      fig.height= 2.9,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, tidyquant, xtable, magrittr, fpp2, ggpubr)
```

# Preface

### Recap of What we Have Covered Last Class

\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Describe AIC, AICc, and BIC and how they are used to measure model fit.}
      \item \textbf{Describe the algorithm used within the auto.arima() function to fit an ARIMA model.}
      \item \textbf{Describe the results of the auto.arima() function.}
	\end{itemize}
\end{block}

### Learning Outcomes for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Describe a seasonal ARIMA model and explain how it applies to a seasonal time series.}
      \item \textbf{Recognize when to fit a seasonal ARIMA model.}
	\end{itemize}
\end{block}



# Seasonal ARIMA Models

### Live Coding: Example 1
```{r birthData}
pacman::p_load(astsa)
birthData = birth # also from the astsa package
```

### In-Class Activity

Use the data [“20 - Netflix_growth_pct_2000.csv”](https://miamioh.instructure.com/courses/142177/files/19770246?module_item_id=2963681). Fit an ARIMA model using the auto.arima() function. Describe the model that is fit, evaluate the model residuals.


### Things to Do to Prepare for Exam (Next Week)

 - Thoroughly read Chapters 6.2-6.8 of our textbook.
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered.  
 
 - No assignment for today's class.
 


---

\maketitle