---
title: "ISA 444: Business Forecasting"
subtitle: "A Self-Paced Review for Weeks 07-08"
author: Fadel M. Megahed
date: 'Spring 2021'
theme: simplex
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
description: >
  Demonstrate your knowledge of the materials covered in class so far. This tutorial is meant to supplement the in-class lectures, graded/non-graded assignments, and everything we have done in class so far. It should also serve as an interactive study guide that can help you evaluate your understanding/knowledge of the material covered so far in class. 
---


## Welcome 

In this tutorial, you will review the concepts of **autocorrelation** and **stationarity**. This tutorial is optional, but if you plan on completing it, you are advised to do the following prior to attempting the tutorial.   

1. Thoroughly went through the slides and WebEx videos for classes 12, 13, and 14.  
2. Read chapters 1, 2, 3.1-3.4, 4 (with the exception of chapter 4.5), and 6.1-6.5 from our textbook.  
3. Be able to dedicate about 30 minutes to go through this entire tutorial.


**For your information, here are all the packages that are loaded with this tutorial.**
```{r setup, include=TRUE, verbose = FALSE, message=FALSE, warning=FALSE}
library(learnr) # used to convert my R Markdown into a tutorial
library(tidyverse)
library(tidyquant)
library(fpp2)
library(magrittr)
library(timetk)
library(plotly)
library(scales)
library(tseries)
tutorial_options(exercise.timelimit = 600) # server sleeps with 10 min of inactivity
```

## Autocorrelation & Stationarity

In class, we have discussed the use of autocorrelation (`acf()` from base R -- i.e., no packages are required) and partial autocorrelation functions (`pacf()` from base R -- i.e., no packages are required) to determine the **type of models** that can be used when modeling different time-series. Below, I provide an overview of what to look for when visualizing each of the `acf()` and `pacf()` functions.   

  - *Does the `acf() %>% autoplot()` die-down or cut-off?* If the acf cuts-off (i.e., it is quickly indicating that as the lags increase, they 'quickly' become not significant), then the time-series being investigated is likely to be stationary. You can/should confirm your insights by running either the `kpss.test()` or the `adf.test()` from the [forecast](https://cran.r-project.org/web/packages/tseries/tseries.pdf) package.
  - *Does the `acf() %>% autoplot()` seem to be geometrically decreasing in magnitude?* If the answer is yes (i.e., the series has an exponential/sinsoidual pattern) and the `pacf() %>% autoplot()` has only lag(1) to be statistically significant (i.e. autocorrelation statistically different than 0)? Then, we have some evidence that an AR(1) may be appropriate to model the time-series.  
  - *Do we have evidence of an MA(n) process?* The acf() is only significant for $n$ lags and it cuts off for the rest of the time-series.

```{r acf, fig.keep='hold', out.width='50%', warning='FALSE', echo=FALSE}
set.seed(2020)
sim1 = arima.sim(model=list(ma=c(.25, .25, .5)),n=10000)
ts.plot(sim1)
acf(sim1, lag.max = 12)
```
  
```{r interpretTS1, echo=FALSE}
question("From the two plots above, what conclusions can you make? [Select all that apply]",
  answer("We have a non-stationary time-series", message = 'The mean seems to be around 0 and the acf cuts off at lag(3) '),
  answer('We have a stationary time-series', correct = TRUE),
  answer("We can model the series using an MA(3) process", correct = TRUE),
  answer('We can model the series using an MA(4) process', message = 'Note that this plot starts at lag(0) and lag(4) is not statistically different from zero'),
  allow_retry = FALSE
)
```

### Statistical Test for Stationarity

Confirm your findings about whether the time-series is stationary or not by building on the code chunk below. 
```{r exercise7, exercise=TRUE, excercise.eval= TRUE}
set.seed(2020)
sim1 = arima.sim(model=list(ma=c(.25, .25, .5)),n=10000)
```


```{r exercise7-solution}
set.seed(2020)
sim1 = arima.sim(model=list(ma=c(.25, .25, .5)),n=10000)

# You can confirm your findings 
adf.test(sim1) # we will reject the null hypothesis of non-stationarity
kpss.test(sim1) # we will not reject the null hypothesis of stationarity
``` 
  