---
title: "ISA 444: Business Forecasting"
subtitle: "01 - Course Overview, Introductions and an Overview of Forecasting"
author: Fadel M. Megahed
date: 'Fall 2020'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, xtable)
```

# Preface 

### Learning Objectives for Today's Class

\begin{block}{\textbf{Main Learning Outcomes}}
	\begin{itemize}
			\item \textbf{Describe course objectives \& structure.}
			\item \textbf{Describe what do we mean by \textcolor{miamired}{forecasting} and explain the \textcolor{miamired}{PIVASE} framework.}
			\item \textbf{Explain the differences between cross sectional, time series, and panel datasets.}
			\item \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
	\end{itemize}
\end{block}


# Course Expectations, Overview & Introductions

### Different Types/Components of Analytics

\begin{block}{\textbf{The analytics journey}}
		\begin{itemize}[<+>]
			\item \textcolor{miamired}{Pre-analytics/Data Management:} where one attempts to extract the needed data for analysis.
			\item \textcolor{miamired}{Descriptive analytics:} where one attempts to understand the data through visualizations and descriptive statistics. 
			\item \textcolor{miamired}{Predictive analytics:} statistical and machine learning models are used.
			\item \textcolor{miamired}{Prescriptive analytics:} mathematical models are used to make recommendations for business actions.
		\end{itemize}
\end{block}


### Miami's Business Analytics Curriculum: A Perspective

\begin{figure}
		\centering
		\adjustbox{max width=\textwidth, frame}{%
			\begin{forest}
				[\LARGE{\textbf{Business Analytics}}
				[\textbf{Data Management}
				[ISA 245]
				[ISA 401]
				[ISA 414]]
				[\textbf{Descriptive}
				[ISA 225]
				[ISA 401]]
				[\textbf{Explanatory/Predictive}
				[ISA 291]
				[\textcolor{miamired}{\textbf{ISA 444}}]
				[ISA 491]]
				[\textbf{Prescriptive}
				[ISA 321]
				[ISA 365]]]
		\end{forest}}
		\caption{My take on the courses within the business analytics major/minor at Miami University.}
	\end{figure}


### Course Objectives

**By the end of this course, you should be able to:**

  - **Explain the purpose of forecasting in a business setting.**  
  - **Use the basic tools of forecasting including plots, summary measures, transformations, measures of forecast accuracy, and prediction intervals.**  
  - **Forecast a nonseasonal time series using simple exponential smoothing.**  
  - **Forecast a nonseasonal time series using linear exponential smoothing.**  
  - **Use decomposition methods and Holt-Winters smoothing methods to forecast a seasonal time series.**   
  - **Use ARIMA models to forecast a time series.**  
  - **Use simple and multiple linear regression models to forecast a time series.**


### Why should you care? - An Exploration of the Job Market {.allowframebreaks}
	
\begin{figure}
	\centering
	\href{https://tinyurl.com/y2pqfgm6}{\includegraphics[width=\textwidth, height = 0.7\textheight, keepaspectratio, frame]{Figures/forecastAnalyst1.png}}
	\caption{Required qualifications for a \textit{forecast analyst} position. Click on the image to open the ad.}
\end{figure}

\begin{figure}
	\centering 
	\href{https://tinyurl.com/y4pau6hk}{\includegraphics[width=\textwidth, height = 0.7\textheight, keepaspectratio, frame]{Figures/forecastAnalyst2.png}}
	\caption{Required qualifications for a \textit{Risk Capacity Planning Analyst} at Paypal. Click on the image to open the ad.}
\end{figure}

\begin{figure}
	\centering 
	\href{https://tinyurl.com/y5v86ga4}{\includegraphics[width=0.7\textwidth, height = 0.6\textheight, keepaspectratio, frame]{Figures/forecastAnalyst3.png}}
	\caption{Required qualifications for a \textit{Lead Business Intelligence Analyst} at Spectrum Health. Click on the image to open the ad.}
\end{figure}

\begin{figure}
	\centering
	\href{https://tinyurl.com/y6l7k2cc}{\includegraphics[width=\textwidth, height = 0.7\textheight, keepaspectratio, frame]{Figures/indeed.png}}
	\caption{Number of full-time, entry-level, \$100,000+ jobs on \textit{Indeed.com}, with the term "forecasting" as of August 14, 2020. Click on the image to update the search.}
\end{figure}


### Instructional Approach

\begin{figure}
		\centering
		\begin{tikzpicture}[node distance=2cm, font=\small]
		\tikzstyle{start} = [rectangle, rounded corners, minimum width=0.3\textwidth, minimum height=1cm,text badly centered, text width=0.3\textwidth, draw=black, thick,font=\normalfont]
		
		\tikzstyle{categories} = [rectangle, rounded corners, minimum width=0.3\textwidth, minimum height=1cm,text centered, text width=0.3\textwidth, draw=black, fill=gray!10, thick]
		
		
		\tikzstyle{arrow} = [->,>= triangle 45]
		
		% Begining of the creation of the figure
		\node (start) [start] {\textbf{Instructional Approach}};
		
		% Level 1 nodes
		\node (ia1) [categories, below of = start, xshift=-0.35\textwidth] {\textbf{\underline{In-class Learning}} \\ (A) Bring your A-game \\ (B) Very active learning env.};
		\node (ia2) [categories, below of = start, xshift = 0in] {\textbf{\underline{Mutual Respect}} \\ (A) Conducive learning env. \\ (B) Interact \& contribute};	
		\node (ia3) [categories, below of = start, xshift=0.35\textwidth] {\textbf{\underline{We will have fun}} \\ (A) Enjoyable class to teach \\ (B) V. cool learning materials};
		
		% Arrows
		\draw [arrow] (start.south) -| (ia1);
		\draw [arrow] (start.south) -| (ia2);
		\draw [arrow] (start.south) -| (ia3);
		\end{tikzpicture}
		\caption{An overview of the instructional approach for ISA 444.}
	\end{figure}


### How will I Evaluate your Learning?

\begin{figure}[h!]
		\centering
		\begin{tikzpicture}[node distance=2cm, font=\small]
		\tikzstyle{start} = [rectangle, rounded corners, minimum width=0.3\textwidth, minimum height=1cm,text centered, text width=0.3\textwidth, draw=black, thick,font=\normalfont]
		
		\tikzstyle{categories} = [rectangle, rounded corners, minimum width=0.2\textwidth, minimum height=1cm,text centered, text width=0.2\textwidth, draw=black, fill=gray!10, thick]
		
		
		\tikzstyle{arrow} = [->,>= triangle 45]
		
		% Begining of the creation of the figure
		\node (start) [start] {\textbf{Evaluation Components}};
		
		% Level 1 nodes
		\node (e1) [categories, below of = start, xshift=-0.35\textwidth] {Non-graded class \\ activities (0\%)};
		\node (e3) [categories, right of = e1, xshift=0.25\textwidth] {Exams \\ (80\%)};
		\node (e2) [categories, right of = e3, xshift=0.25\textwidth] {Graded class \\ activities (20\%)\footnotemark};
		
		\node (e4) [categories, below of = e3, xshift=-0.375\textwidth] {Exam I \\ (20\%)};
		\node (e5) [categories, below of = e3, xshift=-0.125\textwidth] {Exam II \\ (20\%)};
		\node (e6) [categories, below of = e3, xshift=0.125\textwidth] {Exam III \\ (20\%)};
		\node (e7) [categories, below of = e3, xshift=0.375\textwidth] {Exam IV \\ (20\%)};
		
		% Arrows
		\draw [arrow] (start.south) -| (e1);
		\draw [arrow] (start.south) -| (e2);
		\draw [arrow] (start.south) -| (e3);
		
		\draw [arrow] (e3.south) -- (e4.north);
		\draw [arrow] (e3.south) -- (e5.north);
		\draw [arrow] (e3.south) -- (e6.north);
		\draw [arrow] (e3.south) -- (e7.north);
		\end{tikzpicture}
		\caption{An overview of the evaluation components for ISA 444.}
	\end{figure}

\footnotetext{Note that I drop the lowest three graded class assignments/activities. We should have 20+ of these graded assignments/activities during the semester.}


### About Me - My Route to Miami University

\begin{block}{\textbf{Academic Experience}}
		\begin{itemize}
			\item \textbf{Application of Data-Driven Decisions (D\textsuperscript{3})} in \textcolor{miamired}{\textbf{3 Continents}}.
			\item \textcolor{miamired}{\textbf{Interests:}}  Health-care, logistics, occupational safety \& portfolios.
			\item \textcolor{miamired}{Partnered with:} \href{https://www.aflac.com}{Aflac}, \href{https://www.fatiguescience.com/}{Fatigue Science}, \href{https://www.jbhunt.com/
			}{JB Hunt}, \href{https://mavenmachines.com/}{Maven Machines} \& \href{http://tennibot.com/}{Tennibot}
		\end{itemize}
	\end{block}
	\begin{center}
		\fbox{\includegraphics[width=\textwidth, height=0.4\textheight, keepaspectratio]{map.JPG}}\\
		Journey with Data-Driven Decision (D\textsuperscript{3}) Making. \\
	\end{center}


### Your Academic Background Motivation for Taking this Class
\begin{block}{\textbf{In-Class Poll:}}
		\textbf{Please use your phone, computer, or tablet and:}
		\begin{itemize}
			\item Go to \url{https://www.menti.com/}.
			\item Insert the code shown on my screen.
			\item Answer the two questions.
		\end{itemize}
\end{block}


# So What is Forecasting?

### When I Searched for Forecast on the Web
\begin{figure}
  \centering
  \href{https://tinyurl.com/yyktgqz2}{\includegraphics[width=0.9\textwidth, height = 0.7\textheight, frame, keepaspectratio]{Figures/forecast.png}}
  \caption{People tend to link forecasting to the "weather" (at least that is what search engines think).}
\end{figure}

### Definition -- from Bing

\begin{figure}
  \centering
  \href{https://tinyurl.com/y5h5k4aj}{\includegraphics[width=0.9\textwidth, height = 0.7\textheight, frame, keepaspectratio]{Figures/forecastDef.png}}
  \caption{The definition of the term "forecast" as obtained from Bing/Merriam-Webster.}
\end{figure}


### Definition and Purpose

**Forecast** *is a prediction or estimate of an actual outcome expected in a future time period or for another situation.*\footnotemark

- *The purpose of forecasting is to inform the process of planning.*  

- *The purpose of planning is to develop a course of action so that things don't “just continue” based on a no-change forecast.*


\footnotetext{The definition and purpose provided in this slide are from: Ord, K., Fildes, R., \& Kourentzes, N. (2017). Principles of Business Forecasting (2nd ed., p. 3). Wessex Press Inc.}


### Working Definitions -- From Prof. Jones-Farmer

\begin{block}{\textbf{Forecasting}}
\textbf{The process of predicting a future event. The objective of most time series analyses is to \textcolor{miamired}{provide forecasts of future values of the time series.}}
\end{block}

\vspace{\baselineskip}

\begin{block}{\textbf{Time Series}}
A \textbf{\textcolor{miamired}{time series}} is a sequence of observations on a variable measured at successive points in time or over successive periods of time. In the simplest cases, observations are evenly spaced at regular intervals such as hourly, daily, weekly, monthly, or yearly, or at any other regular interval.\footnotemark
\end{block}

\footnotetext{Both definitions are based on \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}


### The FAANG (- AAPL) Time Series Data - an R Exercise [1]

\textcolor{miamired}{Let us get and generate the time series below.} **I promised this class will be fun!!!**

\vspace{-0.5\baselineskip} 

```{r timeseries, echo=FALSE, fig.fullwidth = TRUE}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyquant, tidyverse)
FANG = tq_get(c('FB', 'AMZN', 'NFLX', 'GOOG'),
              from = "2020-01-01")
FANG %>% ggplot(aes(x = date, y = adjusted, group = symbol)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020") +
  facet_wrap(~ symbol, ncol = 2, scale = "free_y") +
  theme_bw()
```

\vspace{-1.25\baselineskip} 

\centering{\textcolor{darkgreen}{COVID-19 did not have any long-lasting effects on Tech stocks.}}


### The FAANG (- AAPL) Time Series Data - an R Exercise [2]

We can actually quantify my statement in \textcolor{darkgreen}{green} from the previous slide by computing the 
percent change in each of the four stock prices when compared to January 2, 2020. \textcolor{miamired}{This can be done as follows: (see live coding session in class).\footnotemark}
```{r pctChanges, echo=FALSE}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyquant, tidyverse)
FANG %>% filter(date == '2020-01-02') %>% select(symbol, adjusted) -> baseFANG
FANG %>% filter(date == '2020-08-12') %>% select(symbol, adjusted) -> currentFANG
colnames(baseFANG)[2] = 'base'
colnames(currentFANG)[2] = 'current'
combinedFANG = merge(baseFANG, currentFANG, by ="symbol") %>% 
  mutate( pctChange = 100*(current - base)/base )
cat(paste0('The percent changes (from January 2, 2020) in the AMZN, FB, GOOG and NFLX',
    '\n', ' stocks are: ', round(combinedFANG[1,4],1), '%, ', 
    round(combinedFANG[2,4],1), '%, ', round(combinedFANG[3,4],1), '%, and ', 
    round(combinedFANG[4,4],1), '%, respectively.'))
```

\footnotetext{The printed numbers from my computations had a current date of August 14, 2020. Thus, the numbers will change (slightly) unless we use the same ending date for our calculations.}


### Why do we Forecast? -- A Conceptual Framework (PIVASE)\footnotemark

- **Purpose:** What do we hope to achieve by generating the forecast? That is, what plans are dependent upon the results of the forecasting exercise? How far ahead do we wish to forecast? We refer to this period as the forecasting horizon.  

- **Information:** What do we know that may help us in forecasting. And when will we know it? Detailed data is only useful if it is available in timely fashion.   

- **Value:** How valuable is the forecast? What would you pay for perfect knowledge?   

- **Analysis:** From analyzing the data can we develop a model that captures its characteristics? And how does it perform on new (hold-out sample) data?   

- **System:** What models and software are needed to meet the needs of the organization?  

- **Evaluation:** How do we know whether a particular forecasting exercise was effective and what the potential is for improvement?

\footnotetext{From: Ord, K., Fildes, R., \& Kourentzes, N. (2017). Principles of Business Forecasting (2nd ed., p. 3-6).}


### Why do Businesses Forecast?

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{Figures/businessForecasts.png}
  \caption{Some of the typical forecasting needs of many organizations.\footnotemark}
\end{figure}

\footnotetext{From: Ord, K., Fildes, R., \& Kourentzes, N. (2017). Principles of Business Forecasting (2nd ed., p. 7).}



# Types of Data Over Time

### Cross Sectional Data [1]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 1: H1B 2020 Data for Senior Data Scientists at Netflix}}\footnotemark

```{r netflixSalaries, results='asis', echo=FALSE}
if(require(pacman)==FALSE) install.packages(pacman)
pacman::p_load(lubridate, rvest, xtable)
read_html("https://h1bdata.info/index.php?em=NETFLIX+INC&job=SENIOR+DATA+SCIENTIST&city=LOS+GATOS&year=2020") %>%
  html_node("#myTable") %>% html_table() -> netflixSalaries
netflixSalaries$`START DATE` = mdy(netflixSalaries$`START DATE`)
netflixSalaries2020 = netflixSalaries %>%  filter(`START DATE` >= '2020-01-01') %>% 
  select(EMPLOYER, `JOB TITLE`, `BASE SALARY`, LOCATION)
print(xtable(netflixSalaries2020, align = c(rep('c', 5)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data scraped from \url{https://h1bdata.info/index.php?em=NETFLIX+INC&job=SENIOR+DATA+SCIENTIST&city=LOS+GATOS&year=2020} on August 14, 2020 using the \href{https://cran.r-project.org/web/packages/rvest/rvest.pdf}{rvest} package in R. The printing was limited to those individuals who started on/after January 01, 2020, with the filters specified in the URL.}


### Cross Sectional Data [2]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 2: Heart Disease Dataset}}\footnotemark

```{r heart, results='asis', echo=FALSE}
set.seed(444)
heart = read.delim("https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/heart/heart.dat",
                  header = FALSE, sep = " ") %>% select(c(1,2,4,8,14)) %>% sample_n(., 13) 
colnames(heart) = c('age', 'sex', 'restingBP', 'maxHR', 'label')
print(xtable(heart, align = c(rep('c', 6)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data sampled from \href{https://archive.ics.uci.edu/ml/datasets/statlog+(heart)}{this UCI Machine Learning Repository}.}


### Cross Sectional Data [3]


**Cross Sectional Data:** Measurements on multiple units, recorded in a single time period.

\textbf{\textcolor{miamired}{Example 3: NBA 2019-2020 Leaders - Top 12 in PTS/Game}}\footnotemark

```{r nba, echo = FALSE, results='asis'}
nbaPerGame = read_html("https://www.basketball-reference.com/leagues/NBA_2020_per_game.html") %>% 
  html_node("#per_game_stats") %>% html_table() %>% filter(!PTS %in% "PTS")
nbaPerGame$PTS = as.numeric(nbaPerGame$PTS)
nbaPerGame %>% arrange(desc(PTS)) %>% select(Player, Pos, Age, Tm, G, FG, `FG%`, `eFG%`, PTS) %>% 
  head(n = 12) -> nbaPerGame
print(xtable(nbaPerGame, align = c('l', 'l', rep('c', 8)) ), comment = FALSE, size = '\\scriptsize', 
      include.rownames=FALSE)
```

\footnotetext{Data scraped from \href{https://www.basketball-reference.com/leagues/NBA_2020_per_game.html}{Basketball-Reference} on August 14, 2020 using the \href{https://cran.r-project.org/web/packages/rvest/rvest.pdf}{rvest} package in R. The printing with limited to the top 12 players and the selected variables.}



### Time Series Data [1]

**Time Series Data:** Comparable measurements recorded on a single (or a few) variables over time (usually a long period of time).

\textbf{\textcolor{miamired}{Example 1: Net Power Generation in the U.S.}}

\begin{figure}
  \centering
  \href{https://www.eia.gov/electricity/data/browser/}{\includegraphics[width=\textwidth, height=0.5\textheight, keepaspectratio, frame]{eiaChart.png}}
  \caption{Net monthly electricity generation in the U.S. from the \href{https://www.eia.gov/electricity/data/browser/}{U.S. Energy Information Adminstration}.}
\end{figure}


### Time Series Data [2]

\textbf{\textcolor{miamired}{Example 2: Stock prices of U.S. Airlines}}

\vspace{-0.5\baselineskip} 

```{r timeseries2, echo=FALSE, fig.fullwidth = FALSE}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyquant, tidyverse)
airlines = tq_get(c('DAL', 'UAL', 'AAL', 'LUV'),
              from = "2020-01-01")
airlines %>% ggplot(aes(x = date, y = adjusted, group = symbol)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020") +
  facet_wrap(~ symbol, ncol = 2, scale = "free_y") +
  theme_bw()
```

\vspace{-1.25\baselineskip} 

\centering{\textcolor{darkgreen}{As of August 14, 2020: COVID-19 had a substantial impact on these airline stocks.}}



### Panel Data


**Panel Data:** Cross sectional measurements (usually many variable) repeated over time (usually over a few time periods).

\textbf{\textcolor{miamired}{Example: World Bank's Data}}\footnotemark

```{r wb, results='asis', echo=FALSE}
pacman::p_load(wbstats)
macroData = wb_data(country = c('CN','EG','US'),
               indicator = c('NY.GDP.MKTP.KD.ZG', 'SP.DYN.LE00.IN', 'SH.DYN.NMRT', 'SH.HIV.INCD.ZS',
                             'SH.MED.BEDS.ZS', 'SH.MED.PHYS.ZS', 'SH.PRV.SMOK', 'SH.STA.TRAF.P5'),
               mrv = 3, return_wide = TRUE, gapfill = TRUE) %>% 
  select(-c(iso2c, country))
print(xtable(macroData, align = c(rep('c', 11)) ), comment = FALSE, size = '\\tiny', 
      include.rownames=FALSE)
```

\footnotetext{Data queried from the \href{https://datacatalog.worldbank.org/}{World Bank Data Catalog} using the \href{https://cran.r-project.org/web/packages/wbstats/wbstats.pdf}{wbstats} package in R. The printed results show a snapshot of 7 variables (out of a much larger panel dataset). You can think of panel data as a cross-sectional dataset with a longitudinal/time component.}




# Components of a Time Series

### Trend [1]

A **trend** is an increasing or decreasing pattern over time.

```{r trendUP, echo=FALSE}
pacman::p_load(tidyquant, tidyverse)
amazon = tq_get(c('AMZN'), from = "2020-04-01", to = "2020-08-14")
amazon %>% ggplot(aes(x = date, y = adjusted)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020", 
       title = "Increasing Trend", 
       subtitle = "The meteoric rise of $AMZN from 2020-04-01 to 2020-08-14") +
  theme_bw()
```


### Trend [2]

A **trend** is an increasing or decreasing pattern over time.

```{r trendDown, echo=FALSE}
pacman::p_load(tidyquant, tidyverse)
delta = tq_get(c('DAL'), from = "2020-02-03", to = "2020-04-30")
delta %>% ggplot(aes(x = date, y = adjusted)) +
  geom_line() + 
  labs(y = "Adjusted Closing Price", x = "2020", 
       title = "Decreasing Trend", 
       subtitle = "The decline in $DAL from 2020-02-03 to 2020-04-30") +
  theme_bw()
```



### Seasonality [1]

**Seasonality** refers to the property of a time series that displays REGULAR patterns that repeat at a constant frequency (*m*).

\begin{figure}
  \centering 
  \includegraphics[width=\textwidth, height = 0.55\textheight, keepaspectratio]{Figures/seasonal.jpg}
  \caption{A time series with a monthly seasonal pattern.\footnotemark}
\end{figure}

\footnotetext{Figure is from \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}


### Seasonality [2]

**Seasonality** refers to the property of a time series that displays REGULAR patterns that repeat at a constant frequency (*m*).

\vspace{-0.5\baselineskip}

```{r retailSales, echo=FALSE}
pacman::p_load(lubridate)
retail = read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1168&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=RSXFSN&scale=left&cosd=1992-01-01&coed=2020-07-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Monthly&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2020-08-15&revision_date=2020-02-01&nd=2010-01-01")
retail %>% ggplot(aes(x = ymd(DATE), y = RSXFSN)) +
  geom_line() + 
  labs(x=NULL,
       title = "Seasonality with an Additive Trend", 
       subtitle = "Retail (- Food Services) from 2010-01-01 to 2020-02-01",
       caption = 'Data from FRED') +
  theme_bw()
```

### Seasonality [3]

```{r airpassengers, echo=FALSE}
pacman::p_load(ggfortify)
data("AirPassengers")
autoplot(AirPassengers) + 
  labs(x=NULL, title = "Seasonality with a Multiplicative Trend", 
       subtitle = "Non-linear trend & seasonal component grows over time",
       caption = 'AirPassengers R Dataset -- Source: Box, G. E. P., Jenkins, G. M. and \n Reinsel, G. C. (1976) Time Series Analysis, Forecasting and Control.') +
  theme_bw()
```


### Cycle [1]
**Cyclical fluctuations** are somewhat irregular (unknown duration).

\vspace{-0.5\baselineskip}

```{r autoSales, echo=FALSE}
pacman::p_load(lubridate)
autoSales = read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1168&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=TOTALSA&scale=left&cosd=1976-01-01&coed=2020-07-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Monthly&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2020-08-15&revision_date=2020-08-15&nd=1976-01-01")
autoSales %>% ggplot(aes(x = ymd(DATE), y = TOTALSA)) +
  geom_line() + 
  labs(x=NULL,
       title = "The cyclical nature of auto sales",
       caption = 'U.S. Bureau of Economic Analysis, Total Vehicle Sales [TOTALSA], retrieved from FRED, Federal Reserve Bank of St. Louis \n https://fred.stlouisfed.org/series/TOTALSA, August 14, 2020.') +
  theme_bw()
```


### Cycle [2]
**Cyclical fluctuations** are somewhat irregular (unknown duration).

\vspace{-0.5\baselineskip}

\begin{figure}
  \includegraphics[width=0.7\textwidth, height = 0.7\textheight, keepaspectratio]{Figures/Step1.jpg}
  \caption{A gait "cycle" from an IMU attached to the ankle. Joint work with the University at Buffalo.}
\end{figure}



# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
	\begin{itemize}
			\item \textbf{Describe course objectives \& structure.}
			\item \textbf{Describe what do we mean by \textcolor{miamired}{forecasting} and explain the \textcolor{miamired}{PIVASE} framework.}
			\item \textbf{Explain the differences between cross sectional, time series, and panel datasets.}
			\item \textbf{Identify and describe the basic components of a time series including trends, seasonal components, and cycles.}
	\end{itemize}
\end{block}


### Things to Do

 - Thoroughly read Chapter 1 of our book, which can be downloaded from the \href{https://wessexlearning.com/collections/business-analytics/products/principles-of-business-forecasting-2nd-ed}{Publisher} (if you have not gotten your book yet).  
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered. 
 
 - Complete the graded assignment (refer to our next slide or Canvas) for more details.
 

### Graded Assignment: Evaluating your Retention/Focus

Please go to \href{https://miamioh.instructure.com/courses/123532/quizzes/317526}{Canvas (click here)} and answer the four questions. **Due August 21, 2020 [11:59 PM,
Ohio local time].** 

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of the material covered in Class 01. In order to prepare for this, you should have either actively attended class and/or watched the recording from WebEx.

**General Guidelines:**  

\vspace{-0.5\baselineskip}

  - Individual assignment.  
  - This is a timed assignment (i.e. once you start the assignment you will have 25 minutes to complete 4 questions). The purpose of the time limit is to help you evaluate how much you have retained/understood from class. If the concepts we covered today are well-understood, this should take 10-15 minutes.  
  - Proctorio is NOT required for this assignment.  
  - You will need to have R installed (or accessible through the \href{https://virtualpc.fsb.miamioh.edu/RDWeb/Pages/en-US/default.aspx}{Remote Desktop})


---

\maketitle